---
title: "Using Large Language Models (LLMs) in Social Science Research via API-Access"
author: "Arthur Mühl"
date: "`r Sys.Date()`"
format:
  html:
    code-fold: show
execute:
  eval: true
  echo: true
  warning: false
  message: false
  freeze: false
engine: knitr
---

```{r echo=FALSE}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
```

# 1. Introduction: Why LLMs Matter in Social Science

Large Language Models (LLMs) represent a paradigm shift in computational social science. Traditional supervised learning approaches often require large, manually annotated datasets, which are costly and time-consuming. LLMs, accessed via APIs, allow researchers to perform complex text analysis tasks with minimal labeled data. This enables faster experimentation, democratizes access to cutting-edge NLP tools, and allows analysis of novel or specialized datasets. 

**Use Case: Stance Detection**  
To illustrate, consider stance detection: identifying whether a social media post supports, opposes, or is neutral toward a policy. LLMs can perform this with minimal examples, which is particularly useful for rapidly evolving datasets.

# 2. Getting Started: Setup and Authentication

## Obtain and Secure API Keys

Accessing LLM APIs requires creating an account with a provider and generating a secret API key.

- **OpenAI**: Create an account at openai.com, navigate to API keys, and generate a new key.
- **Anthropic**: Sign up at anthropic.com, go to “API Keys,” and generate a secret key.
- **Cohere**: Register at cohere.ai, go to API dashboard, and copy your key.

Generating the key and using it accordingly might not function before switching to a paid-plan or adding funds to your account, depending on the provider. You should see the pricing overview of the specific provider before continuing.

### Best Practices for Credential Management

Never hard-code your API key in scripts. Instead, store it as an environment variable. This process differs depending on your operating system.

```{r eval=FALSE}
# Linux/macOS (bash)
Sys.setenv(OPENAI_API_KEY = "your_openai_key_here")

# Windows (PowerShell)
system('setx OPENAI_API_KEY "your_openai_key_here"')
```

Retrieving the key in R:

```{r eval=FALSE}
api_key <- Sys.getenv("OPENAI_API_KEY")
```

## Setting Up the Coding Environment

Install required libraries:

```{r eval=FALSE}
install.packages("httr")     # for API requests
install.packages("jsonlite") # for handling JSON
```

Example initialization for OpenAI:

```{r eval=FALSE}
library(httr)
library(jsonlite)

base_url <- "https://api.openai.com/v1/chat/completions"
headers <- add_headers(
  Authorization = paste("Bearer", api_key),
  `Content-Type` = "application/json"
)
```

Example initialization for Anthropic (alternative provider):

```{r eval=FALSE}
library(httr)
library(jsonlite)

anthropic_base_url <- "https://api.anthropic.com/v1/messages"
headers_anthropic <- add_headers(
  `x-api-key` = api_key,
  `anthropic-version` = "2023-06-01",
  `content-type` = "application/json"
)
```

## First request: a minimal test call

After you have set your API key and base configuration, send a tiny request to verify everything works.

### OpenAI (chat completions)

```{r eval=FALSE}
base_url <- "https://api.openai.com/v1/chat/completions"
headers <- add_headers(
  Authorization = paste("Bearer", api_key),
  `Content-Type` = "application/json"
)
body <- list(
  model = "gpt-3.5-turbo",
  messages = list(list(role = "user", content = "Hello! Please reply with a single short sentence.")),
  max_tokens = 50,
  temperature = 0.2
)
res <- POST(url = base_url, body = toJSON(body, auto_unbox = TRUE), encode = "raw", headers)
stop_for_status(res)
out <- content(res, as = "parsed", type = "application/json")
cat(out$choices[[1]]$message$content)
```

### Anthropic (messages)

```{r eval=FALSE}
library(httr)
library(jsonlite)
anthropic_base_url <- "https://api.anthropic.com/v1/messages"
headers_anthropic <- add_headers(
  `x-api-key` = api_key,
  `anthropic-version` = "2023-06-01",
  `content-type` = "application/json"
)
body <- list(
  model = "claude-3-5-sonnet-20240620",
  max_tokens = 50,
  messages = list(list(role = "user", content = "Hello! Please reply with a single short sentence."))
)
res <- httr::POST(
  url = anthropic_base_url,
  body = jsonlite::toJSON(body, auto_unbox = TRUE),
  encode = "raw",
  headers_anthropic
)
stop_for_status(res)
raw <- content(res, as = "parsed", type = "application/json")
# Anthropic returns content as a list of blocks; print the first text block
cat(raw$content[[1]]$text)
```

# 3. Prompting Strategies: From Basics to Advanced

Prompting transforms an LLM into a task-specific analytical tool. There are several different prompting strategies, that each have their own strengths and weaknesses.

## Zero-Shot Prompting

Provide instructions without examples:

```{r}
prompt <- "Classify the stance of this tweet about climate policy: 'We need stronger climate action now.'"
```

Use when you lack labeled examples. Fast but may be less accurate.

## Few-Shot Prompting

Include labeled examples to guide the model:

```{r eval=FALSE}
prompt <- "
Tweet: 'I support climate regulations.'
Stance: Pro

Tweet: 'Climate policies hurt the economy.'
Stance: Anti

Tweet: 'Renewable energy is the future.'
Stance: Pro

Tweet: 'We need stronger climate action now.'
Stance:"
```

Improves accuracy by providing context.

## Chain-of-Thought Prompting

Chain-of-thought (CoT) prompting guides the model to reason explicitly in steps before producing the final answer. This is useful for multi-hop reasoning, coding tasks, data transformations, and methodological explanations.

### How to structure CoT prompts

- Set the goal and constraints clearly.
- Ask the model to think step by step.
- Specify how to move from step to step (state, check, decide, proceed).
- Instruct the model to output a final concise answer after the steps.
- When needed, request a machine-readable final summary (e.g., JSON) after the reasoning.

For more in-depth information about prompting strategies, see the gesis [gesis prompting guide](https://quarto.org).

## Example for Zero-Shot-Prompting with ChatGPT (OpenAI)

Zero-shot prompts do not include task-specific examples; you describe the task clearly and let the model infer the pattern.

```{r eval=FALSE}
base_url <- "https://api.openai.com/v1/chat/completions"
headers <- add_headers(
  Authorization = paste("Bearer", api_key),
  `Content-Type` = "application/json"
)

zero_shot_prompt <- paste(
  "You are a helpful assistant.",
  "Task: Summarize the following paragraph in one sentence, focusing on the key takeaway.",
  "Paragraph: Open-source vector databases have matured rapidly, offering better recall, faster indexing, and lower operational cost.",
  "Keep it under 25 words.",
  sep = "\n"
)

body <- list(
  model = "gpt-3.5-turbo",
  messages = list(list(role = "user", content = zero_shot_prompt)),
  max_tokens = 80,
  temperature = 0.3
)

res <- POST(url = base_url, body = toJSON(body, auto_unbox = TRUE), encode = "raw", headers)
stop_for_status(res)
out <- content(res, as = "parsed", type = "application/json")
cat(out$choices[[1]]$message$content)
```

## Trade-offs

- **Accuracy vs. Cost**: More complex prompting or longer examples may increase token usage.
- **Stability**: Zero-shot can be unstable; few-shot usually produces more consistent results.

# 4. Structuring Outputs for Research

LLMs often return verbose, unstructured text. For automated research pipelines, structured outputs are essential. Without it, parsing errors and inconsistencies can occur, especially in large-scale analysis.

### Function Calling / JSON Output

We can define a JSON schema to force consistent, machine-readable output:

```{r eval=FALSE}
prompt <- "
Classify the tweet into 'Pro', 'Anti', or 'Neutral'.
Return a JSON object like:
{
  'stance': 'Pro'
}
Tweet: 'We need stronger climate action now.'
"

body <- toJSON(list(
  model = "gpt-4o-mini",
  messages = list(list(role = "user", content = prompt)),
  temperature = 0.0
))

response <- POST(base_url, headers, body = body)
result <- content(response, "parsed")
stance <- fromJSON(result$choices[[1]]$message$content)$stance
```

This ensures outputs integrate seamlessly into analysis workflows, allowing automated aggregation, visualization, and statistical modeling. Check here for more in-depth information about [function calling](https://www.promptingguide.ai/applications/function_calling).

# 5. Practical Challenges in Large-Scale Research

## Costs: Token-Based Pricing and Scalability

LLM APIs charge per token (approx. a word or subword). For instance:

| Provider       | Model          | Input Cost ($/1M tokens)  | Output Cost ($/1M tokens) |
|----------------|----------------|---------------------------|---------------------------|
| OpenAI         | GPT-4o         | 5.00                      | 20.00                     |
| OpenAI         | GPT-4o-mini    | 0.60                      | 2.40                      |
| Anthropic      | Claude 3 Opus  | 15.00                     | 75.00                     |
| Anthropic      | Claude 3 Haiku | 0.25                      | 1.25                      |
| Google         | Gemini 1.5 Pro | 1.25-2.50                 | 5.00-10.00                |
| Cohere         | Command-R+     | 3.00                      | 15.00                     |

Cost control strategies:

- Use smaller models for large corpora.
- Reduce prompt length.
- Reuse few-shot templates instead of regenerating.

## Latency: Efficiency Strategies

Factors affecting latency:

- Model size (larger = slower).
- Prompt length.
- Network speed.

Mitigation:

- Use streaming APIs for incremental output.
- Optimize prompt length.
- Use low-latency models like GPT-4o-mini.

## Hallucinations: Risks and Mitigation

LLMs can generate plausible-sounding but false information. Example:

- Model may claim a social media post references a law that doesn’t exist.

Mitigation:

- Human-in-the-loop review.
- Output filtering.
- Provide detailed and accurate context.

# Using a VM or GPU for LLM workflows in R

Many API-based workflows run fully in the cloud provider, so you do not need a local GPU. However, VMs can help with:

- Large-scale pre/post-processing (embeddings, chunking, retrieval).
- Hosting vector databases and caches.
- Running open-source models locally (when needed).

### Option A: Cloud VM (no GPU) for API orchestration

- Provision a small VM (e.g., 2–4 vCPUs, 8–16 GB RAM).
- Install R, RStudio Server, and system libs.
- Store secrets via environment variables or a secret manager.

Example bootstrap (Ubuntu):

```bash
sudo apt-get update
sudo apt-get install -y r-base gdebi-core
# optional: RStudio Server
wget https://download2.rstudio.org/server/jammy/amd64/rstudio-server-2024.04.2-764-amd64.deb
sudo gdebi -n rstudio-server-2024.04.2-764-amd64.deb
```

In R, manage keys safely:

```{r eval=FALSE}
Sys.setenv(OPENAI_API_KEY = "...", ANTHROPIC_API_KEY = "...")
api_key <- Sys.getenv("OPENAI_API_KEY")
```

### Option B: GPU VM for local/open models

If you need to run open models (e.g., via ollama/text-generation-webui or transformers):

- Choose a GPU VM (e.g., 1x T4/A10/A100 depending on model size).
- Ensure drivers + CUDA/cuDNN are installed or use a prebuilt image.
- Prefer containerized stacks to avoid driver mismatches.

Docker example with ollama server:

```bash
curl -fsSL https://ollama.com/install.sh | sh
ollama pull llama3
ollama run llama3 "Hello"
```

Then call from R via httr to http://localhost:11434.

### Practical tips

- Log prompts, responses, and model versions for reproducibility.
- Use caching for repeated prompts.
- Monitor costs and storage; rotate logs with sensitive data.
